Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2019-04-11 manuscript/homophily_manuscript.tex
Response to 3rd plos review

We believe that most or all earlier studies of gender homophily were
hindered by a largely unacknowledged statistical issue that we will
refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with
the conceptually similar Wahlund effect in population genetics {[}59{]}.
The Wahlund effect makes it deceptively difficult to infer gender-based
preferences simply by counting the number of same- and mixed-gender
coauthorships. Essentially, whenever coauthorship data are sampled from
two or more discrete sets of literature, which vary in the author gender
ratio and which are largely not connected by collaboration, the number
of same-gendered coauthors will be inflated. This can give the
impression that authors preferentially publish with same-gendered
colleagues even if no gender preferences exist, or if the true
preference is for opposite-gendered colleagues (`gender heterophily').
For example, a sample of literature containing a mixture of
bioinformatics and cell biology papers will probably contain an excess
of mostly-male and mostly-female author lists, simply because
researchers usually collaborate within their own discipline, and because
the author gender ratio is more male-biased in bioinformatics than in
cell biology {[}5{]}.

\begin{figure}
\centering
\includegraphics{../figures/Fig1.pdf}
\caption{The Wahlund effect can make it appear as if authors publish
with same-gendered colleagues disproportionately often, even if
collaboration is completely random with respect to gender. Here,
coloured circles represent male and female authors, and coauthors are
linked with lines. Across the whole set of ten papers, there is an
apparent excess of same-gender collaborations: there are six same-gender
papers and only four mixed-gender papers, which is fewer than the
\(10\times2\times0.5\times0.5 = 5\) mixed-gender papers expected under
the null hypothesis that authors assort randomly. However, within each
subset, there is no evidence that authors prefer to publish with
same-gendered individuals (if anything, this small dataset suggests
gender heterophily). The Wahlund effect will tend to inflate the
frequency of same-gender coauthorships whenever the data is composed of
two or more disconnected subsets of literature with different author
gender ratios; these subsets could be research disciplines, older versus
newer papers, or papers from authors in different countries. The example
countries and disciplines were selected based on data in {[}5{]}.
\label{wahlund_plot}}
\end{figure}

In the present study, we test whether life sciences researchers tend to
co-publish with same-gendered colleagues, while controlling for the
Wahlund effect as strictly as possible. We use a recently-published
dataset describing the gender of 35.5 million authors from 9.15 million
articles indexed on PubMed {[}5{]}. Holman et al. {[}5{]} reported large
differences in the gender ratio of authors across research disciplines,
journals, countries, and across the years 2002-2016. We therefore tested
for gender homophily while restricting our analysis to particular
journals (i.e.~research specialties), time periods, and countries. We
quantified gender assortment using a metric called \(\alpha'\) {[}60{]},
which is positive when same-gender authors publish together more often
than expected (gender homophily), negative when opposite-gender authors
publish together more often than expected (heterophily), and equal to
zero when authors assort randomly with respect to gender (see Methods).

@@

\autoref{alpha_histograms} shows the distribution of \(\alpha'\)
estimates in 2015-2016 across all journals for which we recovered
sufficient data, when \(\alpha'\) was calculated for all authors, first
authors only, or last authors only. Most journals had positive values of
\(\alpha'\) (77-92\%, depending on time period and author type; S1
Data), and for many of these the false discovery rate (FDR)-corrected
p-values suggested that \(\alpha'\) was significantly greater than zero
(1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6;
S1 Data). Only 2/2077 journals had statistically significant heterophily
(i.e. \(\alpha' < 0\)) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The
remaining 606 or 787 journals (in 2015 and 2005 respectively) had a
value of \(\alpha'\) not significantly different from zero, consistent
with the null hypothesis of random assortment with respect to gender. We
also confirmed that in most journals (S2 Data) and most research
disciplines (S3 Data, S1 Fig), the majority of papers had multiple
authors.

\begin{figure}
\centering
\includegraphics{../figures/Fig2.pdf}
\caption{Of the 2116 journals for which we had adequate data in
2015-2016, 825 showed statistically significant evidence of gender
homophily (denoted by \(\alpha' > 0\)), and 1 showed statistically
significant evidence of heterophily (\(\alpha' < 0\)), after false
discovery rate correction. In the stacked density plot, the white area
shows the number of journals for which homophily was significantly
stronger than expected under the null hypothesis (corrected p
\textless{} 0.05), while the blue area shows all the remainder. Patterns
were similar whether \(\alpha'\) was calculated for all authors, for
first authors only, or for last authors only. Points in the right panel
show \(\alpha'\) for individual journals.\label{alpha_histograms}}
\end{figure}

@@

\autoref{alpha_histograms} illustrates the variance in journal homophily
values (\(\alpha'\)) across scientific disciplines. All disciplines had
positive mean \(\alpha'\) (averaged over journals), although homophily
appeared somewhat stronger in some disciplines than others (e.g.~mean
\(\alpha'\) was \(0.12{\pm}0.02\) for Urology journals and
\(0.03{\pm}0.01\) for Veterinary Medicine journals;
\autoref{alpha_histograms}, S4 Data). However, there was no formal
evidence for consistent differences in \(\alpha'\) between disciplines:
the random factor `Discipline' explained around 1\% of the variance in
\(\alpha'\) in the two linear mixed models described in the previous
section (see \autoref{alpha_histograms} and mixed models in Online
Supplementary Material). Thus, whatever processes are responsible for
producing positive \(\alpha'\) values appear to be similarly strong in
all the disciplines we examined.

There was no indication that journals publishing on a wide range of
topics have higher \(\alpha'\) values than more specialised journals,
due to the Wahlund effect. For example, the journal category
`Multidisciplinary' -- which includes journals like \emph{PLOS ONE},
\emph{Nature}, \emph{Science}, and \emph{PNAS} -- did not have markedly
elevated \(\alpha'\) (\autoref{alpha_histograms}). This result suggests
that our estimates of homophily, and estimates from some of the earlier
studies listed in the Introduction, are probably not markedly inflated
by the presence of disparate research topics (with variable author
gender ratios) being published within individual journals.

Nevertheless, when we calculated \(\alpha\) across all non-single-author
papers in our entire 15-year PubMed dataset (as before, excluding papers
where at least one author's gender was unknown; n = \textgreater{}3
million papers, \textgreater{}16 million authors), we found that
\(\alpha\) was 0.126. This figure is almost double the median value of
\(\alpha'\) for individual journals (Figure 2; \(\alpha' = 0.070\) for
`All authors'), suggesting that lumping together papers from different
fields and different time periods can indeed produce spurious evidence
for gender homophily as outlined in Figure 1.

@@

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, on
average (\autoref{author_number}; statistical results in Online
Supplementary Material). Papers with 3, 4 or \(\ge5\) authors had
essentially identical average \(\alpha'\) values. The variance in
\(\alpha'\) across journals was also a little higher for 2-author papers
compared to the remainder (\autoref{author_number}), though part of this
variance is due to the reduced sample size (in terms of number of
authors) for the 2-author papers. One possible explanation for this
finding is that 2-authors papers might be more likely to have an author
list that is evenly split between career stages (e.g.~a postgraduate
student and their supervisor), increasing the chance that the authors
are mixed gender (see Figure 6). The result also suggests that the
processes responsible for gender homophily are similar in small
(e.g.~3-author) and larger (5+ author) collaborations (and across
disciplines where small versus large collaborations are the norm).

\begin{figure}
\centering
\includegraphics{../figures/Fig3.pdf}
\caption{The coefficient of homophily (\(\alpha'\)) was slightly less
positive when calculated for two-author papers only, relative to papers
with longer author lists. The individual points, whose distribution is
summarised by the violin plots, correspond to individual journals. The
larger white points show the mean for each group (and its 95\% CIs), as
calculated by a Bayesian meta-regression model accounting for repeated
measures of \(\alpha'\) within journals, as well as the precision with
which \(\alpha'\) was estimated. \label{author_number}}
\end{figure}

@@

We next tested whether researchers are more or less likely to publish
with same-gendered colleagues in strongly gender-biased disciplines
(e.g.~Surgery or Nursing), relative to disciplines with a comparatively
gender-balanced workforce (e.g.~Psychiatry). We found a positive,
non-linear relationship between the overall gender ratio of all authors
publishing in a particular journal {[}5{]}, and the estimated value of
\(\alpha'\) for all authors and for first authors
(\autoref{alpha_gender_ratio}). Journals with a balanced or
female-biased author gender ratio tended to have higher \(\alpha'\)
(i.e.~stronger homophily) than journals with a male-biased author gender
ratio (GAM smooth terms p \textless{} 0.001; Online Supplementary
Material). The relationship was not statistically significant when
\(\alpha'\) was calculated for last authors (GAM, p = 0.142), though the
trend appeared similar (\autoref{alpha_gender_ratio}).

\begin{figure}
\centering
\includegraphics{../figures/Fig4.pdf}
\caption{There is a weakly positive, non-linear relationship between the
gender ratio of authors publishing in a journal, and the coefficient of
homophily (\(\alpha'\)). Specifically, journals with 50\% women authors
or higher tended to have more same-sex coauthorships than did journals
with predominantly men authors. This relationship held whether
\(\alpha'\) was calculated for all authors, first authors only, or last
authors only. A negative value on the x-axis denotes an excess of men
authors, a positive value denotes an excess of women authors, and zero
denotes gender parity (i.e.~equal numbers of men and women). The lines
were fitted using generalised additive models with the smoothing
parameter \(k\) set to 3. \label{alpha_gender_ratio}}
\end{figure}

@@

We observed a noisy but statistically significant linear relationship
between standardised journal impact factor and \(\alpha'\), such that
journals with a high impact factor for their discipline had weaker
gender homophily than did journals with a low impact factor for their
discipline (\autoref{impact_factor}; linear regression: \(R^2\) = 0.043,
\(t_{1415}\) = -8.0, p \textless{} 0.0001). The slope of the regression
was \(-0.012{\pm}0.0015\), indicating that increasing the
discipline-standardised impact factor by one standard deviation is
associated with a reduction in \(\alpha'\) of 0.012.

\begin{figure}
\centering
\includegraphics{../figures/Fig5.pdf}
\caption{Journal impact factor (expressed relative to the average for
the discipline) is negatively correlated with \(\alpha'\). The
relationship is noisy (\(R^2\) = 0.043), but the results suggest that
journals with strong homophily tend to have lower impact factors than
journals with weak homophily in the same discipline.
\label{impact_factor}}
\end{figure}

@@

When we restricted the analysis by country, we observed statistically
significant homophily for 72 of the 325 journal-country combinations
tested (64 unique journals and 18 unique countries), and no significant
heterophily (S4-S5 Fig). Additionally, the values of \(\alpha'\)
calculated for each journal-country combination were only very slightly
lower than the \(\alpha'\) values calculated for the journal as a whole
(i.e.~when pooling papers from different countries, as was done to make
\autoref{alpha_histograms}): on average, the difference in \(\alpha'\)
was only 0.002 (S6 Fig). These results suggest that our findings of
widespread homophily in the main analysis were not driven solely by a
Wahlund effect resulting from gender differences between countries.

@@

Given that we cannot identify individual researchers or their career
stages, we used a simple model to derive the theoretical expectations
for \(\alpha\) when the gender ratio differs between career stages (see
Methods). As shown in \autoref{simulation}, we predict that \(\alpha\)
is expected to be non-zero, even if collaborators are randomly selected
with respect to gender, provided that there is a gender gap between
career stages. The extent to which \(\alpha\) deviates from zero depends
on the relative frequencies of collaboration within and between career
stages (rows and columns in Figure 6), and the size of the gender gap
between stages (x- and y-axes in Figure 6). When \textgreater{}50\% of
coauthor pairs comprise one early-career and one established researcher,
we expect gender heterophily (\(\alpha < 0\)) whenever the gender ratio
differs between career stages. Conversely, when \textgreater{}50\% of
collaborations are between people at the same career stage, we expect
gender homophily (\(\alpha > 0\)). In a few parameter spaces (shown in
red; \autoref{simulation}), \(\alpha\) was quite high, and overlapped
with the values that we estimated (\autoref{alpha_histograms}).

\begin{figure}
\centering
\includegraphics{../figures/Fig6_inkscape.pdf}
\caption{When the gender ratio of early-career researchers is not equal
to the gender ratio among established researchers, the null expectation
for \(\alpha\) is not necessarily zero. Specifically, if most
collaborations occur between career stages, there will be an excess of
mixed-gender collaborations (\(\alpha < 0\), blue areas), while if most
collaborator pairs comprise two people at the same career stage, there
will be an excess of same-gender collaborations (\(\alpha > 0\), red
areas). However, the conditions required for strong gender homophily
(i.e.~the red areas) are quite restrictive, making it unlikely that this
issue can fully explain the homophily observed in our study.
Additionally, in research disciplines where between-career stage
collaboration is common and there is a shortage of women among
established researchers (i.e.~the blue areas), our study will
underestimate the strength of gender homophily. Contour lines denote
increments of 0.1. \label{simulation}}
\end{figure}

Despite this overlap, \autoref{simulation} suggests that our main
conclusions (and those of other studies of gender homophily) are
probably robust to this career stage issue. We only expect strongly
positive \(\alpha\) when A) the gender ratio is highly skewed across
career stages (e.g.~a 5-fold difference), and B) collaborations between
early and established researchers are very rare (e.g. \textless{}10\% of
the total). Both of these conditions seem unlikely to be true for most
fields: the gender gap across careers stages is generally less
pronounced {[}1,5{]}, and it is very common for early-career researchers
to co-publish with an established mentor {[}61{]}. However, one can get
\(\alpha > 0\) for realistic combinations of parameters, e.g.~a moderate
shortage of women in senior positions coupled with a moderate excess of
within-career stage collaboration, suggesting this effect might
contribute to some of the homophily observed by this and previous
studies.

Lastly, we note that if there is a gender gap between career stages and
coauthorships between early-career and established researchers comprise
\textgreater{}50\% of the total, then the baseline expectation for
\(\alpha\) is actually less than zero (blue areas in
\autoref{simulation}). Therefore, it is possible that researchers
preferentially assort with same-gendered collaborators even more
strongly than implied by our results, at least for certain journals or
research disciplines.

@@

We found evidence that researchers work with same-gendered coauthors
more often than expected under the null model, even after implementing
stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our
study therefore reaffirms earlier studies' conclusions {[}49--57,62{]}
using stricter methodology, and generalises their results across the
life sciences. Relatively few journals had \(\alpha'\) values below
zero, and almost no journals showed statistically significant gender
heterophily after controlling for multiple testing. The excess of
same-gender coauthorships was quite large: many journals had
\(\alpha' > 0.1\), indicating that the gender ratio of men's and women's
coauthors differs by \textgreater{}10\% in absolute terms. In relative
terms, our findings are even more striking: for example, if men have
20\% female coauthors and women have 30\% (i.e. \(\alpha' = 0.1\) in a
field with a typical gender ratio {[}5{]}), then women publish with
women 50\% more often than men do.

An important limitation of our study is that we cannot reliably
determine the cause(s) of the observed excess of same-gender
coauthorships. As well as the obvious interpretation -- conscious or
unconscious selection of same-gender collaborators by men, by women, or
both -- our results could be partly explained by uncontrolled Wahlund
effects. However, we suspect the contribution of these uncontrolled
artefacts to be minor, for four reasons: we found positive \(\alpha'\)
after controlling for three obvious sources of Wahlund effect; there was
no inflation of \(\alpha'\) in highly multidisciplinary journals
relative to specialised journals; restricting the data by country
yielded similar estimates of \(\alpha'\); and we used a simple model to
show that differences in gender ratio between career stages are unlikely
to fully explain our results. On balance, we believe the data suggest
that it is likely that some researchers preferentially select
same-gendered collaborators, although it is difficult to ascertain what
proportion of people show such a preference, or how much the strength of
the preference varies between individuals. We also note that even in a
world in which everyone selected their collaborators at random with
respect to gender, a high proportion of individual researchers would
have entirely same-gendered collaborators by chance alone (especially in
gender-biased disciplines); thus, individuals who only have
same-gendered co-authors are not necessarily doing anything differently
from people with gender-balanced co-authors.

We hypothesised that disciplines with a strongly skewed gender ratio
might show the strongest gender homophily, e.g.~because being in the
minority might increase people's motivation to seek out same-gendered
colleagues. Contrary to this hypothesis, we found no evidence that
gender homophily is restricted to particular disciplines: \(\alpha'\)
was similarly high across the board (\autoref{alpha_histograms}).
Interestingly, gender homophily was weakest for journals with a
male-biased author gender ratio, and strongest in journals with a
female-biased author gender ratio. One possible reason is that men are
more likely to preferentially seek out male collaborators in fields
where men are a minority, relative to the homophily displayed by women
in fields where women are a minority. However, this latter result only
has tentative statistical support since our sample contains few journals
in which most authors are women (\autoref{alpha_gender_ratio}).

We also found that gender homophily was marginally stronger in 2015-2016
relative to 2005-2006. Although this trend might reflect a change in the
gender preferences of researchers seeking collaborators, there are
alternative (and perhaps more likely) explanations. For example, this
trend might result from the increasing number of women working in senior
positions in STEMM over the past decade {[}63--65{]}. As shown in
\autoref{simulation}, if enough coauthorships are between junior and
senior researchers, a large gender gap between career stages can give
the appearance of heterophily. As this gender gap between career stages
lessens, the observed values of \(\alpha'\) may increase.

>>>>

We believe that most or all earlier studies of gender homophily were
hindered by a largely unacknowledged statistical issue that we will
refer to as the Wahlund effect (Fig 1), by analogy with the conceptually
similar Wahlund effect in population genetics {[}59{]}. The Wahlund
effect makes it deceptively difficult to test for gender-based co-author
choice simply by counting the relative number of same- and mixed-gender
coauthorships. Essentially, the Wahlund effect means that whenever
coauthorship data are sampled from two or more discrete sets of
literature, which vary in the author gender ratio and which are largely
unconnected by collaboration, the number of same-gendered coauthors will
be inflated. This can give the impression that authors preferentially
publish with same-gendered colleagues even if no gender preferences
exist, or if the true preference is for opposite-gendered colleagues
(`gender heterophily'). For example, a sample of literature containing a
mixture of bioinformatics and cell biology papers will probably contain
an excess of mostly-male and mostly-female author lists, simply because
researchers usually collaborate within their own discipline, and because
the author gender ratio is more male-biased in bioinformatics than in
cell biology {[}5{]}.

\vspace{2cm}

\textbf{Fig 1}: The Wahlund effect can make it appear as if authors
publish with same-gendered colleagues disproportionately often, even if
collaboration is completely random with respect to gender. Here,
coloured circles represent male and female authors, and coauthors are
linked with lines. Across the whole set of ten papers, there is an
apparent excess of same-gender collaborations: there are six same-gender
papers and only four mixed-gender papers, which is fewer than the
\(10\times2\times0.5\times0.5 = 5\) mixed-gender papers expected under
the null hypothesis that authors assort randomly. However, within each
subset, there is no evidence that authors prefer to publish with
same-gendered individuals (if anything, this small dataset suggests
gender heterophily). The Wahlund effect will tend to inflate the
frequency of same-gender coauthorships whenever the data is composed of
two or more disconnected subsets of literature with different author
gender ratios; these subsets could be research disciplines, older versus
newer papers, or papers from authors based in different countries. The
example countries and disciplines were selected based on data in
{[}5{]}. \vspace{2cm}

In the present study, we test whether life sciences researchers tend to
co-publish with same-gendered colleagues, while controlling for the
Wahlund effect as strictly as we are able. We use a recently-published
dataset describing the gender of 35.5 million authors from 9.15 million
articles indexed on PubMed {[}5{]}. Holman et al. {[}5{]} reported large
differences in the gender ratio of authors across research disciplines,
journals, countries, and across the years 2002-2016. We therefore tested
for gender homophily while restricting our analysis to particular
journals (a proxy for research specialties), time periods, and
countries. We quantified gender assortment using a metric called
\(\alpha'\) {[}60{]}, which is positive when same-gender authors publish
together more often than expected (gender homophily), negative when
opposite-gender authors publish together more often than expected
(heterophily), and equal to zero when authors assort randomly with
respect to gender (see Methods).

@@

Fig 2 shows the distribution of \(\alpha'\) estimates in 2015-2016
across all journals for which we recovered sufficient data, when
\(\alpha'\) was calculated for all authors, first authors only, or last
authors only. Most journals had positive values of \(\alpha'\) (77-92\%,
depending on time period and author type; S1 Data), and for many of
these the false discovery rate (FDR)-corrected p-values suggested that
\(\alpha'\) was significantly greater than zero (1469/2077 journals were
significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077
journals had statistically significant heterophily (i.e.
\(\alpha' < 0\)) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The
remaining 606 or 787 journals (in 2015 and 2005 respectively) had a
value of \(\alpha'\) not significantly different from zero, consistent
with the null hypothesis of random assortment with respect to gender. We
also confirmed that in most journals (S2 Data) and most research
disciplines (S3 Data, S1 Fig), the majority of papers had multiple
authors.

\vspace{2cm}

\textbf{Fig 2}: Of the 2116 journals for which we had adequate data in
2015-2016, 825 showed statistically significant evidence of gender
homophily (denoted by \(\alpha' > 0\)), and 1 showed statistically
significant evidence of heterophily (\(\alpha' < 0\)), after false
discovery rate correction. In the stacked density plot, the white area
shows the number of journals for which homophily was significantly
stronger than expected under the null hypothesis (corrected p
\textless{} 0.05), while the blue area shows all the remainder. Patterns
were similar whether \(\alpha'\) was calculated for all authors, for
first authors only, or for last authors only. Points in the right panel
show \(\alpha'\) for individual journals. \vspace{2cm}

@@

Fig 2 illustrates the variance in journal homophily values (\(\alpha'\))
across scientific disciplines. All disciplines had positive mean
\(\alpha'\) (averaged over journals), although homophily appeared
somewhat stronger in some disciplines than others (e.g.~mean \(\alpha'\)
was \(0.12{\pm}0.02\) for Urology journals and \(0.03{\pm}0.01\) for
Veterinary Medicine journals; Fig 2, S4 Data). However, there was no
formal evidence for consistent differences in \(\alpha'\) between
disciplines: the random factor `Discipline' explained around 1\% of the
variance in \(\alpha'\) in the two linear mixed models described in the
previous section (see Fig 2 and mixed models in Online Supplementary
Material). Thus, the causal mechanisms underlying the observed positive
\(\alpha'\) values appear to be similarly strong in all the disciplines
we examined.

There was no indication that journals publishing on a wide range of
topics have higher \(\alpha'\) values than more specialised journals due
to the Wahlund effect (Fig 1). For example, the journal category
`Multidisciplinary' -- which includes general interest journals like
\emph{PLOS ONE}, \emph{Nature}, \emph{Science}, and \emph{PNAS} -- did
not have markedly elevated \(\alpha'\) (Fig 2). This result suggests
that our estimates of homophily, and estimates from some of the earlier
studies of homophily listed in the Introduction, are probably not
markedly inflated by the presence of disparate research topics (with
variable author gender ratios) being published within individual
journals.

Nevertheless, when we calculated \(\alpha\) across all non-single-author
papers in our entire 15-year PubMed dataset (as before, excluding papers
where at least one author's gender was unknown; n = \textgreater{}3
million papers, \textgreater{}16 million authors), we found that
\(\alpha\) was 0.126. This figure is almost double the median value of
\(\alpha'\) for individual journals (Fig 2; \(\alpha' = 0.070\) for `All
authors'), suggesting that lumping together papers from different fields
and different time periods can indeed produce spurious evidence for
gender homophily as outlined in Fig 1.

@@

Papers with two authors had significantly lower (but still positive)
\(\alpha'\) values relative to papers with more than two authors, while
papers with 3, 4 or \(\ge5\) authors had essentially identical average
\(\alpha'\) values (Fig 3). Specifically, the posterior estimate of mean
\(\alpha'\) was 0.014 (95\% CIs: 0.002 - 0.026) for 2-author papers and
0.065 (95\% CIs: 0.056 - 0.074) for 3-author papers (and roughly the
same for 4- and \(\ge5\)-author papers; Fig 3). One possible explanation
for this finding is that 2-authors papers are more likely to have an
author list that is evenly split between career stages (e.g.~a
postgraduate student and their supervisor), increasing the chance that
the authors are mixed gender (see section `Theoretical expectations for
\(\alpha\) when the gender ratio differs between career stages'). The
result also suggests that the causal mechanisms responsible for gender
homophily are similar in small (e.g.~3-author) and larger (\(\ge5\)
author) collaborations (and across disciplines where small versus large
collaborations are the norm).

\vspace{2cm}

\textbf{Fig 3}: The coefficient of homophily (\(\alpha'\)) was slightly
less positive when calculated for two-author papers only, relative to
papers with longer author lists. The individual points, whose
distribution is summarised by the violin plots, correspond to individual
journals. The larger white points show the mean for each group (and its
95\% CIs), as calculated by a Bayesian meta-regression model accounting
for repeated measures of \(\alpha'\) within journals, as well as the
precision with which \(\alpha'\) was estimated. \vspace{2cm}

@@

We next tested whether researchers are more or less likely to publish
with same-gendered colleagues in strongly gender-biased disciplines
(e.g.~Surgery or Nursing), relative to disciplines with a comparatively
gender-balanced workforce (e.g.~Psychiatry). We found a positive,
non-linear relationship between the gender ratio across all the authors
publishing in a particular journal {[}5{]}, and the estimated value of
\(\alpha'\) for all authors and for first authors, but not last authors
(S4 Fig). Journals with a balanced or female-biased author gender ratio
tended to have higher \(\alpha'\) (i.e.~stronger homophily) than
journals with a male-biased author gender ratio (GAM smooth term p =
0.0002 for all-author homophily, p \textless{} 0.0001 for first-author
homophily, and p = 0.13 for last-author homophily).

@@

We observed a noisy but statistically significant linear relationship
between standardised journal impact factor and \(\alpha'\), such that
journals with a high impact factor for their discipline had weaker
gender homophily than did journals with a low impact factor for their
discipline (Fig 4; linear regression: \(R^2\) = 0.043, \(t_{1415}\) =
-8.0, p \textless{} 0.0001). The slope of the regression was
\(-0.012{\pm}0.0015\), indicating that increasing the
discipline-standardised impact factor by one standard deviation is
associated with a reduction in \(\alpha'\) of 0.012. The Spearman
correlation coefficient was -0.19 (p \textless{} 0.0001).

\vspace{2cm}

\textbf{Fig 4}: Journal impact factor (expressed relative to the average
for the discipline) is negatively correlated with \(\alpha'\). The
relationship is noisy (\(R^2\) = 0.043, Spearman correlation = -0.19),
but the results indicate that journals with strong gender homophily tend
to have lower impact factors than journals from the same discipline that
have weaker homophily. \vspace{2cm}

@@

When we restricted the analysis by country, we observed statistically
significant homophily for 72 of the 325 journal-country combinations
tested (64 unique journals and 18 unique countries), and no significant
heterophily (S5-S6 Fig). Additionally, the values of \(\alpha'\)
calculated for each journal-country combination were only very slightly
lower than the \(\alpha'\) values calculated for the journal as a whole
(i.e.~when pooling papers from different countries, as was done to make
Fig 2): on average, the difference in \(\alpha'\) was only 0.002 (S7
Fig). These results suggest that our findings of widespread homophily in
the main analysis were not driven solely by a Wahlund effect resulting
from gender differences between countries.

@@

Given that we cannot identify individual researchers or their career
stages, we used a simple model to derive the theoretical expectations
for \(\alpha\) when the gender ratio differs between career stages (see
Methods). As shown in Fig 5, we predict that \(\alpha\) is expected to
be non-zero, even if collaborators are randomly selected with respect to
gender, provided that there is a gender gap between career stages. The
extent to which \(\alpha\) deviates from zero depends on the relative
frequencies of collaboration within and between career stages (rows and
columns in Fig 5), and the size of the gender gap between stages (\(x\)-
and \(y\)-axes in Fig 5). When \textgreater{}50\% of coauthor pairs
comprise one early-career and one established researcher, we expect
gender heterophily (\(\alpha < 0\)) whenever the gender ratio differs
between career stages. Conversely, when \textgreater{}50\% of
collaborations are between people at the same career stage, we expect
gender homophily (\(\alpha > 0\)). In a few parameter spaces (shown in
red; Fig 5), \(\alpha\) was quite high, and overlapped with the values
that we estimated (Fig 2).

\vspace{2cm}

\textbf{Fig 5}: When the gender ratio of early-career researchers is not
equal to the gender ratio among established researchers, the null
expectation for \(\alpha\) is not necessarily zero. Specifically, if
most collaborations occur between career stages, there will be an excess
of mixed-gender collaborations (\(\alpha < 0\), blue areas), while if
most collaborator pairs comprise two people at the same career stage,
there will be an excess of same-gender collaborations (\(\alpha > 0\),
red areas). However, the conditions required for strong gender homophily
(i.e.~the red areas) are quite restrictive, making it unlikely that this
issue can fully explain the homophily observed in our study.
Additionally, in research disciplines where between-career stage
collaboration is common and there is a shortage of women among
established researchers (i.e.~the blue areas), our study will
underestimate the strength of gender homophily. Contour lines mark
increments of 0.1. \vspace{2cm}

Despite this overlap, Fig 5 suggests that our main conclusions (and
those of other studies of gender homophily) are probably robust to this
career stage issue. We only expect strongly positive \(\alpha\) when A)
the gender ratio is highly skewed across career stages (e.g.~a 5-fold
difference), and B) collaborations between early and established
researchers are very rare (e.g. \textless{}10\% of the total). Both of
these conditions seem unlikely to be true for most fields: the gender
gap across careers stages is generally less pronounced {[}1,5{]}, and it
is very common for early-career researchers to co-publish with an
established mentor {[}61{]}. However, one can get \(\alpha > 0\) for
realistic combinations of parameters, e.g.~a moderate shortage of women
in senior positions coupled with a moderate excess of within-career
stage collaboration, suggesting this effect might contribute to some of
the homophily observed by this and previous studies.

Lastly, we note that if there is a gender gap between career stages and
coauthorships between early-career and established researchers comprise
\textgreater{}50\% of the total, then the baseline expectation for
\(\alpha\) is actually less than zero (blue areas in Fig 5). Therefore,
it is possible that researchers preferentially assort with same-gendered
collaborators even more strongly than implied by our results, at least
for certain journals or research disciplines.

@@

We found evidence that researchers work with same-gendered coauthors
more often than expected under the null model, even after implementing
stringent controls for Wahlund effects of the kind illustrated in Fig 1.
Our study therefore reaffirms earlier studies' conclusions
{[}49--57,62{]} using stricter methodology, and generalises their
results across the life sciences. Relatively few journals had
\(\alpha'\) values below zero, and almost no journals showed
statistically significant gender heterophily after controlling for
multiple testing. The excess of same-gender coauthorships was quite
large: many journals had \(\alpha' > 0.1\), indicating that the gender
ratio of men's and women's coauthors differs by \textgreater{}10\% in
absolute terms. In relative terms, our findings are even more striking:
for example, if men have 20\% female coauthors and women have 30\% (i.e.
\(\alpha' = 0.1\) in a field with a typical gender ratio {[}5{]}), then
women publish with women 50\% more often than men do.

An important limitation of our study is that we cannot reliably
determine the cause(s) of the observed excess of same-gender
coauthorships. As well as the obvious interpretation -- conscious or
unconscious selection of same-gendered collaborators by men, women, or
both genders -- our results could be partly explained by uncontrolled
Wahlund effects. However, we suspect the contribution of these
uncontrolled artefacts to be minor, for four reasons: we found positive
\(\alpha'\) after controlling for three obvious sources of Wahlund
effect; there was no inflation of \(\alpha'\) in highly
multidisciplinary journals relative to specialised journals; restricting
the data by country yielded similar estimates of \(\alpha'\); and our
modelling work suggested that differences in gender ratio between career
stages are unlikely to fully explain our results. On balance, we believe
the data suggest that it is likely that some researchers preferentially
select same-gendered collaborators, although it is difficult to
ascertain what proportion of people show such a preference, or how much
the strength of the preference varies between individual researchers. We
also note that even in a world in which everyone selected their
collaborators at random with respect to gender, a high proportion of
individual researchers would have entirely same-gendered collaborators
by chance alone (especially in gender-biased disciplines); thus,
individuals who only have same-gendered co-authors are not necessarily
doing anything differently from people with gender-balanced co-authors.

We hypothesised that disciplines with a strongly skewed gender ratio
might show the strongest gender homophily, e.g.~because being in the
minority might increase one's motivation to seek out same-gendered
colleagues. Contrary to this hypothesis, we found no evidence that
gender homophily is restricted to particular disciplines: \(\alpha'\)
was similarly high across the board (Fig 2). Interestingly, gender
homophily was weakest for journals with a male-biased author gender
ratio, and strongest in journals with a female-biased author gender
ratio. One possible reason is that men are more likely to preferentially
seek out male collaborators in fields where men are a minority, relative
to the homophily displayed by women in fields where women are a
minority. However, this latter result only has tentative statistical
support since our sample contains few journals in which most authors are
women (S4 Fig).

We also found that gender homophily was marginally stronger in 2015-2016
relative to 2005-2006. Although this trend might reflect a change in the
gender preferences of researchers seeking collaborators, there are
alternative (and perhaps more likely) explanations. For example, this
trend might result from the increasing number of women working in senior
positions in STEMM over the past decade {[}63--65{]}. As shown in Fig 5,
if enough coauthorships are between junior and senior researchers, a
large gender gap between career stages can give the appearance of
heterophily. As this gender gap between career stages lessens, the
observed values of \(\alpha'\) may increase.

Regarding our finding of weaker homophily among 2-author papers, we
suspect that many 2-author teams comprise a student/postdoc and a senior
staff member, making these teams especially likely to be mixed-gender,
due to the greater shortage of women among senior researchers {[}1,5{]}.
Assuming this interpretation is correct, this result suggests that our
reported \(\alpha'\) values may underestimate the strength of peoples'
preferences for same-gendered collaborators; essentially, women seeking
a senior collaborator could be constrained to work mostly with men,
meaning that people's ideal and realised gender preferences would be
mismatched. On a related note, Ghiasi et al. {[}51{]} argue that women
in engineering are ``compliant {[}in reproducing{]} male-dominated
scientific structures'' because they do not collaborate often enough
with other women (for reference, Figure 7 in {[}51{]} implies that
coauthorships involving two women are \(c\). 30\% more frequent than
expected under random assortment). By contrast, we feel that it may be
counter-productive to recommend that women collaborate primarily with
other women, e.g.~because this constrains women's options (particularly
in fields like the one studied by Ghiasi et al. -- engineering -- where
90\% of professors are men {[}1{]}). Instead, we suggest that
researchers of both genders can help to close the gender gap in STEMM.
In the context of collaboration, one way to do this is to undertake
self-examination to ensure that one is not inadvertently overlooking or
excluding women among potential students and colleagues. One should also
take care to treat male and female collaborators equally, e.g.~in terms
of training and mentoring, allocation of work, and how one descibes the
collaboration to other people (e.g.~in conference presentations, on the
lab website, or in the `Author contributions' section of a paper).
Experimental work suggests that unconscious bias causes people to
undervalue women's research achievements {[}20{]}, and a study of author
contribution statements found observational evidence that menial or
under-valued tasks are more often assigned to women while more
prestigious tasks are assigned to men {[}61{]}.
